Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2023-12-15 main.tex
:pencil: Add MIMO PC SV toy example

\usepackage[table]{xcolor}
\usepackage{adjustbox}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{booktabs}
\usepackage[makeroom]{cancel}
\usepackage[american]{circuitikz}
\usepackage{cite}
\usepackage{fixmath}
\usepackage[acronym]{glossaries-extra}
\usepackage{hyperref}
\usepackage{import}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage[short]{optidef}
\usepackage{pgfplots}
\usepackage{ragged2e}
\usepackage[subtle]{savetrees}
\usepackage{siunitx}
\usepackage{stfloats}
\usepackage[caption=false,font=footnotesize,subrefformat=parens,labelformat=parens]{subfig}
\usepackage{tabularx}
\usepackage{tikz}

@@

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For general optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below, where $f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)$ is the objective function evaluated over $\bigl\{ \{\mathbf{\Theta}_{g'}^{(r+1)}\}_{g'<g}, \{\mathbf{\Theta}_{g'}^{(r)}\}_{g'>g} \bigr\}$.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				{\nabla_g^\mathrm{E}}^{(r)} = \frac{\partial f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)}{\partial \mathbf{\Theta}_g^*};
				\label{eq:rcg_gradient_euclidean}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				{\nabla_g^\mathrm{R}}^{(r)} = {\nabla_g^\mathrm{E}}^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {{\nabla_g^\mathrm{E}}^{(r)}}^\mathsf{H};
				\label{eq:rcg_gradient_riemannian}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl({\nabla_g^\mathrm{R}}^{(r)} - {\nabla_g^\mathrm{R}}^{(r-1)}\bigr) {{\nabla_g^\mathrm{R}}^{(r)}}^\mathsf{H}\Bigr)}{\mathrm{tr}\Bigl({\nabla_g^\mathrm{R}}^{(r-1)} {{\nabla_g^\mathrm{R}}^{(r-1)}}^\mathsf{H}\Bigr)};
				\label{eq:rcg_weight}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				\mathbf{D}_g^{(r)} = {\nabla_g^\mathrm{R}}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)};
				\label{eq:rcg_direction}
			\end{equation}
			\item Determine the Armijo step size\footnote{To double the step size, simply square the argument instead of recomputing the matrix exponential, i.e., $\exp(2 \mu_g \mathbf{D}_g) = \exp^2(\mu_g \mathbf{D}_g)$.}
			\begin{equation}
				\mu_g^{(r)} = \arg\max_{\mu_g} f\Bigl(\exp\bigl(\mu_g \mathbf{D}_g^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr);
				\label{eq:rcg_step}
			\end{equation}
			\item Perform rotational update along local geodesics
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \exp\Bigl(\mu_g^{(r)} \mathbf{D}_g^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}.
				\label{eq:rcg_update}
			\end{equation}
		\end{enumerate}

@@

		Fig.~\ref{sm:pc_rate_sx} illustrates how \gls{ris} configuration influences the \gls{mimo} \gls{pc} achievable rate.
		To ensure a \qty{20}{bit/s/Hz} transmission, an \gls{snr} of \qty{13.5}{\dB} is required for a 8T4R system.
		This value decreases to \qty{12.5}{\dB} (resp. \qty{8}{\dB}) when 32- (resp. 256-) element diagonal \gls{ris} is present.
		If tetrads can be formed in \gls{bd} \gls{ris}, the \gls{snr} can be reduced by another \qty{20}{\percent} (resp. \qty{44}{\percent}).
		Further increase in $L$ yields a marginal gain and incurs quadratic number of connections.
		Thus, dyadic or tetradic \gls{bd} \gls{ris} usually strike a good balance between performance and complexity.

	\end{subsection}
\end{section}

>>>>

\usepackage[table]{xcolor}
\usepackage{adjustbox}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{booktabs}
\usepackage[makeroom]{cancel}
\usepackage[american]{circuitikz}
\usepackage{cite}
\usepackage{fixmath}
\usepackage[acronym]{glossaries-extra}
\usepackage{hyperref}
\usepackage{import}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage[short]{optidef}
\usepackage{pgfplots}
\usepackage{ragged2e}
\usepackage[subtle]{savetrees}
\usepackage{siunitx}
\usepackage{stfloats}
\usepackage[caption=false,font=footnotesize,subrefformat=parens,labelformat=parens]{subfig}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{epstopdf}

@@

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For general optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below, where $f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)$ is the objective function also evaluated over $\bigl\{ \{\mathbf{\Theta}_{g'}^{(r+1)}\}_{g'<g}, \{\mathbf{\Theta}_{g'}^{(r)}\}_{g'>g} \bigr\}$.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				{\nabla_g^\mathrm{E}}^{(r)} = \frac{\partial f\bigl(\mathbf{\Theta}_g^{(r)}\bigr)}{\partial \mathbf{\Theta}_g^*};
				\label{eq:rcg_gradient_euclidean}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				{\nabla_g^\mathrm{R}}^{(r)} = {\nabla_g^\mathrm{E}}^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {{\nabla_g^\mathrm{E}}^{(r)}}^\mathsf{H};
				\label{eq:rcg_gradient_riemannian}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				\gamma_g^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl({\nabla_g^\mathrm{R}}^{(r)} - {\nabla_g^\mathrm{R}}^{(r-1)}\bigr) {{\nabla_g^\mathrm{R}}^{(r)}}^\mathsf{H}\Bigr)}{\mathrm{tr}\Bigl({\nabla_g^\mathrm{R}}^{(r-1)} {{\nabla_g^\mathrm{R}}^{(r-1)}}^\mathsf{H}\Bigr)};
				\label{eq:rcg_weight}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				\mathbf{D}_g^{(r)} = {\nabla_g^\mathrm{R}}^{(r)} + \gamma_g^{(r)} \mathbf{D}_g^{(r-1)};
				\label{eq:rcg_direction}
			\end{equation}
			\item Determine the Armijo step size\footnote{To double the step size, simply square the argument instead of recomputing the matrix exponential, i.e., $\exp(2 \mu_g \mathbf{D}_g) = \exp^2(\mu_g \mathbf{D}_g)$.}
			\begin{equation}
				\mu_g^{(r)} = \arg\max_{\mu_g} f\Bigl(\exp\bigl(\mu_g \mathbf{D}_g^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr);
				\label{eq:rcg_step}
			\end{equation}
			\item Perform rotational update along local geodesics
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \exp\Bigl(\mu_g^{(r)} \mathbf{D}_g^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}.
				\label{eq:rcg_update}
			\end{equation}
		\end{enumerate}

@@

		Fig.~\ref{sm:pc_rate_sx} illustrates how \gls{ris} configuration influences the \gls{mimo} \gls{pc} achievable rate.
		To ensure a \qty{20}{bit/s/Hz} transmission, an \gls{snr} of \qty{13.5}{\dB} is required for a 8T4R system.
		This value decreases to \qty{12.5}{\dB} (resp. \qty{8}{\dB}) when 32- (resp. 256-) element diagonal \gls{ris} is present.
		If tetrads can be formed in \gls{bd} \gls{ris}, the \gls{snr} can be reduced by another \qty{20}{\percent} (resp. \qty{44}{\percent}).
		Further increase in $L$ yields a marginal gain and incurs $\mathcal{O}(L^2)$ connections.
		We thus conclude dyadic or tetradic \gls{bd} \gls{ris} usually strike a good balance between performance and complexity.
	\end{subsection}

	\begin{subsection}{Channel Singular Value Redistribution}
		In this subsection, we first show the channel shaping advantage of \gls{bd} \gls{ris} by a toy example, then analyzes this \emph{capability} under specific setups.
		Finally, we characterize the \emph{Pareto front} of channel singular values via optimization approach.

		\begin{figure}
			\centering
			\includegraphics[width=\columnwidth]{assets/simulation/pc_singular_toy.eps}
			\caption{Channel singular value shaping by diagonal and symmetry unitary \gls{ris} for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ without direct link.}
			\label{sm:pc_singular_toy}
		\end{figure}

		Consider a toy example with $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (2, 2, 2)$ and assume the direct link is absent.
		The diagonal \gls{ris} is $\mathbf{\Theta}^\mathrm{D} = \mathrm{diag}(e^{\jmath \theta_1}, e^{\jmath \theta_2})$ while the unitary \gls{ris} has 4 independent angular parameters
		\begin{equation}
			% \mathbf{\Theta}^\mathrm{D} =
			% \begin{bmatrix}
			% 	e^{\jmath \theta_1} & 0              \\
			% 	0              & e^{\jmath \theta_2}
			% \end{bmatrix}, \quad
			\mathbf{\Theta}^\mathrm{U} = e^{\jmath \phi} \begin{bmatrix}
				e^{\jmath \alpha} \cos \psi  & e^{\jmath \beta} \sin \psi   \\
				-e^{-\jmath \beta} \sin \psi & e^{-\jmath \alpha} \cos \psi
			\end{bmatrix}.
		\end{equation}
		When the direct link is absent, $\phi$ has no impact on the singular value because $\mathrm{sv}(e^{\jmath \phi} \mathbf{A}) = \mathrm{sv}(\mathbf{A})$.
		For a fair comparison, we enforce \emph{symmetry} with $\beta = \pi / 2$.
		Fig.~\ref{sm:pc_singular_toy} illustrates all possible channel singular values achieved by diagonal and symmetry unitary \gls{ris}.
		Clearly, \gls{bd} \gls{ris} provides a much wider dynamic range of $\sigma_1(\mathbf{H})$ and $\sigma_2(\mathbf{H})$ than diagonal \gls{ris}, despite using the same number of elements and parameters.
		A larger gap is expected when the symmetry constraint can be relaxed.

		\begin{subsubsection}{Rank-1 Indirect Channel}
			% When the \gls{ris} is in the vicinity of the transmitter / receiver, the forward / backward channel is often rank-1 and so is the indirect channel.
			% This is because
			When the \gls{ris} is in the vicinity of the transmitter / receiver, the forward / backward channel is often rank-1.
			The indirect channel is also rank-1 because
			\begin{equation}
				\mathrm{rank}( \mathbf{H}^\mathrm{B} \mathbf{\Theta} \mathbf{H}^\mathrm{F} ) \le \min \bigl( \mathrm{rank}(\mathbf{H}^\mathrm{B}), \mathrm{rank}(\mathbf{\Theta}), \mathrm{rank}(\mathbf{H}^\mathrm{F}) \bigr).
			\end{equation}
			% That is to say, \gls{bd} \gls{ris}
		\end{subsubsection}
	\end{subsection}
\end{section}
